# Unit 3 Notes | Statistics for Business II

## Differences of Two Proportions

- Variances of independent variables **add**

$$SE_{\hat{p}_1 - \hat{p}_2} = \sqrt{SE_{\hat{p}_1}^2 + SE_{\hat{p}_2}^2} = \sqrt{\dfrac{p_1 (1-p_1)}{n_1} + \dfrac{p_2 (1-p_2)}{n_2}}$$

- Even if you're subtracting two noisy measurements, the **uncertainty** from both still adds up
  - Think of it like using two shaky rulers: subtracting doesn’t cancel the shakiness — it just combines it

```{r}

set.seed(1)       # ensures reproducibility
B <- 1000          # number of simulations
n <- 500           # sample size per group
p1 <- 0.5          # true proportion in group 1
p2 <- 0.4          # true proportion in group 2

# Create empty vectors to store simulated means and SDs
mean_x1 <- mean_x2 <- numeric(B)
sd_x1 <- sd_x2 <- numeric(B)

# Loop to simulate samples for both groups
for (i in 1:B) {
  
  # Generate random binary outcomes for group 1 (success = 1, failure = 0)
  x1 <- rbinom(n, size = 1, prob = p1)
  mean_x1[i] <- mean(x1)  # sample proportion for group 1
  sd_x1[i]   <- sd(x1)    # sample SD for group 1
  
  # Repeat for group 2
  x2 <- rbinom(n, size = 1, prob = p2)
  mean_x2[i] <- mean(x2)
  sd_x2[i]   <- sd(x2)
  
}

# Theoretical standard error for a sample proportion
sqrt(p1*(1-p1)/n)

# Empirical standard error from simulated data
mean(sd_x1) / sqrt(n)

# Compute 95% confidence bounds around the sampling mean
ub_x1 <- mean(mean_x1) + 1.96 * sqrt(p1*(1-p1)/n)
lb_x1 <- mean(mean_x1) - 1.96 * sqrt(p1*(1-p1)/n)

# Plot histogram of simulated sample proportions
par(mar = c(4.5, 4.5, 1, 1))
hist(mean_x1, breaks = 12, 
     main = "",
     xlab = "Sample Proportion (Group 1)",
     cex.axis = 1.25, cex.lab = 1.5)
abline(v = c(lb_x1, ub_x1), lty = 2, lwd = 2)  # add CI bounds

# Compute simulated differences
diff <- mean_x1 - mean_x2

# Theoretical SE for the difference in proportions
sqrt(p1*(1-p1)/n + p2*(1-p2)/n)

# Empirical SE estimate (not exact but illustrative)
mean(sd_x1 + sd_x2) / sqrt(n + n)

# Compute 95% CIs using correct (additive) and incorrect (subtractive) formulas
## correct
ub_diff_1 <- mean(diff) + 1.96 * sqrt(p1*(1-p1)/n + p2*(1-p2)/n)
lb_diff_1 <- mean(diff) - 1.96 * sqrt(p1*(1-p1)/n + p2*(1-p2)/n)

## incorrect
ub_diff_2 <- mean(diff) + 1.96 * sqrt(p1*(1-p1)/n - p2*(1-p2)/n)
lb_diff_2 <- mean(diff) - 1.96 * sqrt(p1*(1-p1)/n - p2*(1-p2)/n)

# Visualize sampling distribution of the difference
hist(diff, breaks = 12, 
     main = "",
     xlab = "Difference in Sample Proportions (p1 - p2)",
     cex.axis = 1.25, cex.lab = 1.5)

# Add both sets of CI bounds
abline(v = c(lb_diff_1, ub_diff_1), lty = 2, lwd = 2)
abline(v = c(lb_diff_2, ub_diff_2), lty = 3, lwd = 2, col = "red")

# legend
legend("topright",
       legend = c("SE1 + SE2 (correct)", "SE1 - SE2 (incorrect)"),
       lwd = 2, lty = c(2, 3),
       col = c("black", "red"),
       bty = "n")

# Check coverage rates for both formulas
mean(ifelse(diff >= ub_diff_1 | diff <= lb_diff_1, 1, 0))
mean(ifelse(diff >= ub_diff_2 | diff <= lb_diff_2, 1, 0))

```

### Differences of Two Proportions: Example 1

- Study of **ICU patients admitted after a heart attack**
- Randomly assigned:
  - **Treatment:** blood thinner  
  - **Control:** no blood thinner
- Outcome measured: **24-hour survival**

```{r}

# read in data
x <- read.csv("data/cpr.csv")

# convert to matrix
z <- as.data.frame.matrix(table(x))
z$total <- rowSums(z)
z <- rbind(z, total = colSums(z))
z

## calculate n for each group
nt <- z$total[2]
nc <- z$total[1]

## calculate p(survive) for each group
pt <- z$survived[2] / nt
pc <- z$survived[1] / nc

## calculate test statistics
point_est <- pt - pc
se <- sqrt((pt * (1 - pt) / nt) + (pc * (1 - pc) / nc))

z <- data.frame(
  sig_level = c(0.01, 0.05, 0.1),
  z_score = c(2.576, 1.96, 1.645)
)

z$min <- point_est - z$z_score * se
z$max <- point_est + z$z_score * se
z

par(mar = c(4.5, 4.5, 1, 1))
plot(1:3, rep(point_est, 3), 
     ylim = c(min(z$min), max(z$max)),
     xlab = "Significance Level (α)",
     ylab = "Est. with Conf. Int.",
     cex.axis = 1.25,
     cex.lab = 1.5,
     xaxt = "n")
axis(1, at = 1:3, labels = c(z$sig_level), cex.axis = 1.25)
abline(h = 0, lty = 2, lwd = 2)
arrows(1:3, z$min, 1:3, z$max, length = 0.05, angle = 90, code = 3)

```

- We are **90% confident** that blood thinners change the 24-hour survival rate by between -3 and 29 percentage points for patients similar to those in the study

- Because **0% is within this range**, the evidence is inconclusive
  - We cannot determine whether blood thinners help or harm heart attack patients who have undergone CPR

### Differences of Two Proportions: Example 2

- - **5-year clinical trial:** Do fish oil supplements reduce heart attack risk?
- Participants randomly assigned:
  - **Fish oil**
  - **Placebo**
- Compare **heart attack rates** across groups

```{r}

# read in data
x <- read.csv("data/fish_oil.csv")

# convert to matrix
z <- as.data.frame.matrix(table(x))
z$total <- rowSums(z)
z <- rbind(z, total = colSums(z))
z

## calculate n for each group
nt <- z$total[1]
nc <- z$total[2]

## calculate p(hospital) for each group
pt <- z$hospital[1] / nt
pc <- z$hospital[2] / nc

## calculate test statistics
point_est <- pt - pc
se <- sqrt((pt * (1 - pt) / nt) + (pc * (1 - pc) / nc))

z <- data.frame(
  sig_level = c(0.01, 0.05, 0.1),
  z_score = c(2.576, 1.96, 1.645)
)

z$min <- point_est - z$z_score * se
z$max <- point_est + z$z_score * se

# convert to percentage points
point_est <- point_est * 100
z[,3:4] <- z[,3:4] * 100

par(mar = c(4.5, 4.5, 1, 1))
plot(1:3, rep(point_est, 3), 
     ylim = c(min(z$min), 0),
     xlab = "Significance Level (α)",
     ylab = "Est. with Conf. Int. (Pct. Points)",
     cex.axis = 1.25,
     cex.lab = 1.5,
     xaxt = "n")
axis(1, at = 1:3, labels = c(z$sig_level), cex.axis = 1.25)
abline(h = 0, lty = 2, lwd = 2)
arrows(1:3, z$min, 1:3, z$max, length = 0.05, angle = 90, code = 3)

```

- **Estimated effect:** –0.0043  
  - Heart attacks were **0.43 percentage points lower** with fish oil

- **90% confidence interval:** –0.66 to –0.19 percentage points
  - Because **0 is not in the interval**, the reduction is **statistically significant**

### Practical vs. Statistical Significance

- The effect is **statistically significant**, but **very small**:  
  - about **0.4 fewer heart attacks per 100 people**.

- In large samples, even **tiny effects can be statistically significant**.

- **Practical question:** Is such a small reduction worth the **costs, side effects, or effort of treatment?**

```{r}

# effect size relative to mean
## a 27.5% reduction!
point_est / pc

```

### Differences of Two Proportions: Example 3

- A drone quadcopter company is considering a new manufacturer for rotor blades
  - The new manufacturer is more expensive but claims that their higher-quality blades are **3% more reliable**

- Set up the appropriate hypotheses for this test:
  - $H_0: p_{\text{highQ}} - p_{\text{standard}} = 0.03$
  - $H_A: p_{\text{highQ}} - p_{\text{standard}} \neq 0.03$ 

- Using these data, evaluate the hypotheses above at a significance level of 5%.

### Compute the Point Estimate and Standard Error

```{r}
#| echo: true
#| warning: false
#| message: false

x <- read.csv("data/quad.csv")

p_us <- sum(ifelse(x$prospective_supplier == "pass", 1, 0)) / 1000
p_them <- sum(ifelse(x$current_supplier == "pass", 1, 0)) / 1000

point_est <- p_us - p_them
se <- sqrt( p_us * (1 - p_us) / 1000 +  p_them * (1 - p_them) / 1000 )

z <- data.frame(
  sig_level = c(0.01, 0.05, 0.10),
  z_score = c(2.576, 1.96, 1.645)
)

z$min <- (point_est - z$z_score * se)
z$max <- (point_est + z$z_score * se)

# plot estimates with conf. intervals
par(mar = c(4.5, 4.5, 1, 1))
plot(1:3, rep(point_est, 3), 
     ylim = c(min(z$min), max(z$max)),
     xlab = "Significance Level (α)",
     ylab = "Est. with Conf. Int.",
     cex.axis = 1.25,
     cex.lab = 1.5,
     xaxt = "n")
axis(1, at = 1:3, labels = c(z$sig_level), cex.axis = 1.25)
abline(h = 0.03, lty = 2, lwd = 2)
arrows(1:3, z$min, 1:3, z$max, length = 0.05, angle = 90, code = 3)

## z-score
z <- (point_est - 0.03) / se

# simulate null hypothesis
set.seed(1)
sim <- rnorm(10000, mean = 0.03, sd = se)

# Probability of observing a value this extreme or larger
1 - mean(ifelse(point_est >= sim, 1, 0))

# p-value (right-tailed)
1 - pnorm(z)

# visualize the result
par(mar = c(4.5, 4.5, 1, 1))
hist(sim, main = "", xlab = "Simulated Estimates",
     cex.lab = 1.5, cex.axis = 1.25)
abline(v = point_est, lty = 2)
legend("topleft",
       legend = c("Point Est."),
       lty = 2, cex = 1.25, bty = "n")

```

- From the standard normal distribution:
  - The right-tail area is approximately 0.005
  - Doubling for a two-tailed test gives p = 0.01
  - Since p = 0.01 < 0.05, we **reject the null hypothesis**

- We find statistically significant evidence that the higher-quality blades have a pass rate greater than 3% higher than the standard blades — exceeding the company’s claims.

## t-distributions

- When the population standard deviation ($\sigma$) is unknown (which is most of the time), we rely on the $t$-distribution instead of the normal model.
  - The $t$-test helps us quantify uncertainty and make inferences about population parameters using sample data

- We use a **t-distribution with (df = n - 1)** to model the sample mean when the sample size is small.
  - As the number of observations increases, the **degrees of freedom (df)** also increase.
  - With larger df, the **t-distribution more closely resembles the standard normal distribution**.
  
```{r}

# generate 200 observations between -4 and 4
x <- seq(-4, 4, length = 200)

## plot parameters
dfs <- c(8, 4, 2, 1)
cols <- colorRampPalette(c("darkblue", "skyblue"))(4)
par(mar = c(2.5, 4.5, 1, 1))

## plot loop
for (j in seq_along(dfs)) {
  
  df <- dfs[j]
  y <- dt(x, df = df)
  
  if (j == 1) {
    plot(x, y, type = "l", ylab = "Density", xlab = "",
         cex.axis = 1.25, cex.lab = 1.5, lwd = 2, col = cols[j])
  }else {
    lines(x, y, lwd = 2, col = cols[j])
  }
}

## lines and legend on plot
lines(x, dnorm(x), lwd = 2)
legend("topright", legend = c("normal", paste0("df = ", dfs)),
       col = c("black", cols), lwd = 2, bty = "n", cex = 1.25)

# for degrees of freedom 5:100, calculate the critical value
## cv = 2.5% of values fall to the right of this number
## cv = 1.96 with a normal distribution

x <- data.frame(
  df = 5:100,
  qt = qt(.975, df = 5:100)
)

# plot
par(mar = c(4.5, 4.5, 1, 1))
plot(x$df, x$qt,
     type = "l",
     lwd = 4,
     xlab = "Degrees of Freedom",
     ylab = "Critical Value from t-distribution",
     cex.lab = 1.5,
     cex.axis = 1.25,
     ylim = c(1.95, max(x$qt)))
abline(h = 1.96, lty = 2, lwd = 2)

```

### t-statistic equation

$$
t = \dfrac{x_i - \bar{x}}{ \sigma_x / \sqrt{n}}
$$

```{r}

# few number of observations; many simulations
n <- 5
Nsim <- 10000

# prep loop
rejections_normal <- 0
rejections_t <- 0
t <- list()

# loop
for(i in 1:Nsim){
  
  # simulate data (n = 2)
  set.seed(i)
  x <- rnorm(n)
  
  # calc. t-stat
  tstat <- mean(x) / (sd(x)/sqrt(n))
  
  # normal critical value
  if(abs(tstat) > 1.96){
    rejections_normal <- rejections_normal + 1
  }
  
  # correct t critical value
  if(abs(tstat) > qt(0.975, df = n-1)){
    rejections_t <- rejections_t + 1
  }
  
  # store t-stat
  t[i] <- tstat
  
}

# gather data from loop; convert to dist.
t <- density(unlist(t))

# generate a normal dist.
set.seed(1)
z <- rnorm(10000)

# plot
par(mar = c(4.5, 4.5, 1, 1))
plot(density(z), 
     xlim = c(-4, 4),
     lwd = 2,
     xlab = "Test Statistics", 
     ylab = "Density",
     main = "",
     cex.lab = 1.5, 
     cex.axis = 1.25)
lines(t, lty = 2, lwd = 2)
legend("topleft", 
       legend = c("normal", "t (df = 4)"),
       lwd = 2,
       lty = c(1, 2),
       bty = "n",
       cex = 1.25)

# p-value if we use cv = 1.96 (wrong)
rejections_normal / Nsim

# p-value if we use cv = t* (correct)
rejections_t / Nsim

```

### t-distibutions: Example 1

- Let’s apply the t-distribution in a real-world example involving mercury content in dolphin muscle
  - $H_0: Hg = 2$
  - $H_A: Hg > 2$

```{r}

# read in data
y <- read.csv("data/dolphins.csv")

# see data
head(y)

# n?
n <- nrow(y)
n

# prep simulation
Nsim <- 100000
t <- list()

# point est
se_obs <- sd(y$mercury) / sqrt(n)
t_stat_obs <- (mean(y$mercury) - 2) / se_obs
t_stat_obs

# 5% of value fall to the right of this value
crit_value <- qt(0.95, df = n - 1)

# simulate null hypothesis
for(i in 1:Nsim){
  
  # generate data
  set.seed(i)
  x <- rnorm(n, mean = 2, sd = sd(y$mercury))
  
  # summary stats
  point_est <- mean(x)
  se <- sd(x) / sqrt(n)
  
  # t-stat
  t[i] <- (point_est - 2) / se
  
}
t <- unlist(t)

# plot
par(mar = c(4.5, 4.5, 1, 1))
plot(density(t), 
     lwd = 2,
     cex.lab = 1.5, 
     cex.axis = 1.25,
     xlab = "t-stats | Null is true",
     main = "",
     xlim = c(-5, 5))
abline(v = t_stat_obs, lty = 2, lwd = 2)
abline(v = crit_value, lty = 3, lwd = 2, col = "darkgrey")
legend("topleft",
       legend = c("Crit. Value", 
                  "Point Est."),
       lty = c(3, 2),
       col = c("darkgrey", "black"),
       lwd = 2,
       bty = "n",
       cex = 1.25)

# simuated p-value 4 / 100k
table(ifelse(t_stat_obs < t, 1, 0))

# t-score (exp = 0.5)
mean(ifelse(crit_value < t, 1, 0))

# lower bound of conf. int.
se <- sd(y$mercury) / sqrt(n)
mean(y$mercury) - crit_value * se

```

### t-distibutions: Example 2

- Data from Cherry Blossom 10-Mile Run, held each spring in Washington, D.C.
  - In 2006, the average race time was **93.29 minutes**

- Are runners getting faster or slower over time?  
  - $H_0: H_t = 93.29$
  - $H_A: H_t \neq 93.29$

```{r}

# read in data
x <- read.csv("data/dc_race_2017.csv")

# summary stats
summary(x$race_time)
sd(x$race_time)

# test statistic
t_stat <- (mean(x$race_time) - 93.29) / (sd(x$race_time) / sqrt(nrow(x)))

# critical value
crit_value <- qt(0.975, df = nrow(x) - 1)

# generate n = 100; normal dist.
set.seed(1)
n <- 100
df <- n - 1
x <- rnorm(n, mean = 0, sd = 1)
dx <- density(x)

# plot null
par(mar = c(2, 4.5, 1, 1))
plot(dx$x, dx$y, type = "l", lwd = 2, 
     ylab = "Density", cex.lab = 1.5, cex.axis = 1.25)
abline(v = t_stat, lty = 2, lwd = 2)
legend("topleft",
       legend = c("t-statistic"),
       lwd = 2, lty = 2, 
       bty = "n", cex = 1.25)

# plot p-value
x_fill <- dx$x[dx$x >= t_stat]
y_fill <- dx$y[dx$x >= t_stat]
polygon(c(min(x_fill), x_fill, t_stat),
        c(0, y_fill, 0),
        col = rgb(0, 0, 1, 0.3), border = NA)

# simulated p-value; reject the null
1 - mean(ifelse(t_stat > x, 1, 0))

# formula p-value; reject the null
1 - pt(t_stat, df = df)

```

### t-distibutions: Example 3

- Does treatment using embryonic stem cells (ESCs) help improve heart function after a heart attack?
  - Each sheep was **randomly** assigned to either an **ESC treatment group** or a **control group**.  
  - Researchers measured the **change in heart pumping capacity** after treatment.  
  - Positive values indicates stronger recovery.  
  
```{r}

# read in data
y <- read.csv("data/stem_cell.csv")
y

# calculate differences
y$diff <- y$after - y$before

# calculate average differences (mean, sd, and n)
x <- aggregate(diff ~ trmt, y, 
               function(x) c(mean = mean(x),
                             sd = sd(x),
                             n = length(x)))
x <- as.data.frame(do.call(cbind, x))
x[,2:4] <- apply(x[,2:4], 2, as.numeric)
x

# point estimate
mean_diff <- x$mean[2] - x$mean[1]
mean_diff

# standard error
se_c <- x$sd[1]^2 / x$n[1]
se_t <- x$sd[2]^2 / x$n[2]
se <- sqrt(se_c + se_t)
se

# t-statistic
t_stat <- mean_diff / se
t_stat

# p-value
1 - pt(t_stat, df = 8)

```

### t-distibutions: Example 4

- Data: random sample of **150** mothers and their newborns in North Carolina
  - **weight** — the newborn’s birth weight
  - **smoke** — whether the mother **smoked during pregnancy**  
  - $H_0: H_S = H_{NS}$
  - $H_A: H_S \neq H_{NS}$

```{r}

# read in data
y <- read.csv("data/ncbirths.csv")

# randomize
set.seed(1)
y$random_number <- rnorm(nrow(y), mean = 0, sd = 1)
y <- y[order(y$random_number),]

# split
t <- y[y$habit == "smoker",]
c <- y[y$habit == "nonsmoker",]

# combine
y <- rbind(t[1:50,],
           c[1:100,])

# calculate mean, sd and n for babies by mother's smoking
x <- aggregate(weight ~ habit, y, 
               function(x) c(mean = mean(x),
                             sd = sd(x),
                             n = length(x)))
x <- as.data.frame(do.call(cbind, x))
x[,2:4] <- apply(x[,2:4], 2, as.numeric)
x

# point esimtate
mean_diff <- x$mean[x$habit == "smoker"] - x$mean[x$habit == "nonsmoker"]
mean_diff

# standard errors
se_c <- x$sd[x$habit == "nonsmoker"]^2 / x$n[x$habit == "nonsmoker"]
se_t <- x$sd[x$habit == "smoker"]^2 / x$n[x$habit == "smoker"]
se <- sqrt(se_c + se_t)
se

# t-stat
t_stat <- mean_diff / se
t_stat

# p-value
pt(t_stat, df = min(x$n))
```


## Chi-Squared Distributions

- $\chi$ = the Greek letter for "chi" (pronounced like "kai")

- The $\chi^2$ distribution is a continuous probability distribution
  - If a variable $Z$ has the standard normal distribution, then $Z^2$ has the $\chi^2$ distribution with one **degree of freedom**
  
- If $Z_1, Z_2,..., Z_k$ are independent standard normal variables, then...
  - $Z_1^2 + Z_2^2 +...+ Z_k^2$
  - ...has a $\chi^2$ distribution with $k$ **degrees of freedom**
  
```{r}

# generate 10k random, standardized numbers
set.seed(1)
x <- rnorm(10000, mean = 0, sd = 1)
x2 <- x^2

# Two plots side by side
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 2, 1))

# plot 1
hist(x,
     main = "Standard Normal Distribution",
     xlab = "Z")

# plot 2
par(mar = c(4.5, 4.5, 1, 1))
hist(x2,
     main = "Chi-Squared Distribution",
     xlab = "Z^2")

# Reset layout after plotting
par(mfrow = c(1, 1))

```

### Degrees of Freedom

- **Degrees of freedom (df):** number of **independent pieces of information** used to estimate a statistic.
  - $df \neq n$
  - Imagine tossing a coin 100 times -- if you get 55 heads, then you know you got 45 tails (1 df)
  - Imagine a traffic light -- if not yellow or red, then green (2 df)
  
- Calculate the standard deviation of three exam grades
  - $s = \sqrt{\frac{ \Sigma (x_i - \bar{x}) }{n - 1} }$
  - If the mean is already known, then you don't **need** all of the other values in the sample
  - If $\bar{x} = 85$ and $x_1 = 80$ and $x_2 = 85$, then $x_3 = 90$ is no longer independent
  
- **Why it matters**
  - Determines the shape of distributions like **$t$, $F$, and $\chi^2$**.
  - **More df - more information - distributions become narrower and more normal.**
  
### Chi-Squared Distributions: Properties

- mean: $\mu = k$
- variance: $\sigma^2 = 2k$
- mode occurs at $\mu - 2$

```{r}

# expected = 1
mean(x2)

# expected = 2
var(x2)

# estimate chi-squared
k <- 4
for(i in 1:k){
  
  set.seed(i)
  x <- rnorm(10000, mean = 0, sd = 1)
  x2 <- x^2
  
  if(i == 1){
    
    z2 <- x2
    
  }else{
    
    z2 <- z2 + x2
    
  }
}
mean(z2); k
var(z2); 2*k

# simulate data again
## increase variance for better data visualization
set.seed(1)
z <- rnorm(10000, mean = 0, sd = 2)
z2 <- z*z

# generate density plots
y <- data.frame(
  x2 = z2,
  df3 = dchisq(z2, df = 3),
  df6 = dchisq(z2, df = 6),
  df12 = dchisq(z2, df = 12)
)
y <- y[order(y$x2),]

# plot
par(mar = c(4.5, 4.5, 1, 1))
plot(y$x2, y$df3, type = "l",
     main = "",
     xlab = "Chi-Squared Values",
     ylab = "Density",
     cex.axis = 1.25,
     cex.lab = 1.5,
     lwd = 4,
     xlim = c(0, 25))
lines(y$x2, y$df6, lty = 2, col = "blue", lwd = 4)
lines(y$x2, y$df12, lty = 3, col = "red", lwd = 4)
legend("topright",
       legend = c("df = 3",
                  "df = 6",
                  "df = 12"),
       col = c("black", "blue", "red"),
       lty = 1:3, lwd = 4,
       bty = "n", cex = 1.5)
segments(3 - 2, 0, 3 - 2, max(y$df3), lty = 2, lwd = 4)
segments(6 - 2, 0, 6 - 2, max(y$df6), lty = 2, lwd = 4, col = "blue")
segments(12 - 2, 0, 12 - 2, max(y$df12), lty = 2, lwd = 4, col = "red")

```

### Chi-Squared Tests

- This technique is commonly used in two settings:
  - To determine whether a **sample is representative** of a larger population across several groups
  - To evaluate whether observed data **follow a specified theoretical distribution** (e.g., normal, geometric, Poisson)

- Allows us to examine **all bins simultaneously**, rather than testing one or two categories in isolation

$$
\chi^2 = \sum_i \frac{(O_i - E_i)^2}{E_i}
$$

- Where:
  - $O_i$ = observed count in category $i$  
  - $E_i$ = expected count under the null hypothesis

- This statistic follows a **chi-squared distribution** with $(k - 1)$ degrees of freedom, where $k$ is the number of categories

### Chi-Squared Tests: Example 1

- Consider data from a **random sample of 275 jurors** in a small county
  - We would like to test whether this sample is **racially representative** of the eligible population of jurors (i.e., registered voters)
  - $H_0$: Jurors are randomly selected; differences reflect natural sampling variation
  - $H_A$: Jurors are not randomly selected; differences are too large to attribute to chance alone

```{r}

# data
x <- read.csv("data/jurors.csv")
y <- read.csv("data/county_voters.csv")

# summary stats and parameters
e <- table(y$race) / nrow(y) * nrow(x)
o <- table(x$race)
n <- nrow(x)
k <- 4

# test statistic
X2_obs <- sum( (o - e)^2  / e )
X2_obs

# simulate the null
set.seed(1)
B <- 10000

# random multinominal matrix
Y <- t(rmultinom(B, size = n, prob = e))
m <- matrix(e, B, k, byrow=TRUE)
colnames(m) <- colnames(Y) <- names(e)

# simulated
head(Y)

# expected
head(m)

# simulated test statistic
X2_sim <- rowSums( (Y - m)^2 / m )
p_value1 <- mean(X2_sim >= X2_obs)
p_value1

# plot test statistic on chi-squared dist.
par(mar = c(4.5, 4.5, 1, 1))
d_X2_sim <- density(X2_sim)
par(mar = c(4.5, 4.5, 1, 1))
plot(d_X2_sim$x, d_X2_sim$y, type = "l",
     xlab = "Chi-Squared", ylab = "Density",
     cex.lab = 1.5, cex.axis = 1.25, lwd = 2)
abline(v = X2_obs, lty = 2, lwd = 4)

x_tail <- d_X2_sim$x[d_X2_sim$x >= X2_obs]
y_tail <- d_X2_sim$y[d_X2_sim$x >= X2_obs]

polygon(c(X2_obs, x_tail, max(x_tail)),
        c(0, y_tail, 0),
        col = rgb(0,0,1,0.3), border = NA)

text(X2_obs + 1, max(d_X2_sim$y) * 0.4,
     paste0("p ≈ ", round(p_value1, 3)),
     pos = 4, cex = 1.2)
legend("topright",
       legend = c("Point Estimate"),
       lty = 2, lwd = 2, bty = "n",
       cex = 1.25)

```

### Chi-Squared Tests: Example 2

- Suppose we analyze **daily stock returns from the S&P 500** over a 10-year period to test whether market movements are **independent across days**  
  - "Does knowing whether the market went up or down yesterday help predict what happens today?"

- If daily market movements are truly **independent**, the waiting time between positive days should follow a **geometric distribution**

- The geometric distribution gives the probability of waiting $k$ trials to observe the first success
  - $p$ = probability that a given day is an Up day  
  - $X$ = number of days until the next Up day

$$P(X = k) = (1 - p)^{k - 1}p$$

```{r}

# read in "data"
x <- data.frame(
  days = 1:7,
  observed = c(717, 369, 155, 69, 28, 14, 10)
)

# "up" 54.5% of days
p <- 0.545
x$expected <- round((1-p)^((x$days)-1)*p*sum(x$observed)) # geo. distribution
x

# plot
par(mar = c(4.5, 4.5, 1, 1))
plot(x$days, x$observed, type = "l", lwd = 4,
     ylab = "Observed", xlab = "Days",
     cex.axis = 1.25, cex.lab = 1.5)
lines(x$days, x$expected, lty = 2, lwd = 4, col = "blue")
legend("topright", 
       legend = c("Observed", "Expected"),
       lwd = 2, lty = c(1, 2),
       col = c("black", "blue"),
       bty = "n", cex = 1.25)

# test statistic
x2 <- sum( (x$observed - x$expected)^2 / x$expected )
x2

# simulate null
for(i in 1:6){
  set.seed(i)
  z <- rnorm(sum(x$observed), mean = 0, sd = 1)
  if(i == 1){
    y <- z^2
  }else{
    y <- y + z^2
  }
}

# simulated p-value
mean(ifelse(x2 > y, 1, 0))

# formula p-value
pchisq(x2, df = 6)

# plot null
d_X2_sim <- density(y)
pvalue2 <- pchisq(x2, df = 6)
par(mar = c(4.5, 4.5, 1, 1))
plot(d_X2_sim$x, d_X2_sim$y, type = "l",
     xlab = "Chi-Squared", ylab = "Density",
     cex.lab = 1.5, cex.axis = 1.25, lwd = 2)
abline(v = x2, lty = 2, lwd = 4)

x_tail <- d_X2_sim$x[d_X2_sim$x >= x2]
y_tail <- d_X2_sim$y[d_X2_sim$x >= x2]

polygon(c(X2_obs, x_tail, max(x_tail)),
        c(0, y_tail, 0),
        col = rgb(0,0,1,0.3), border = NA)

text(X2_obs + 1, max(d_X2_sim$y) * 0.4,
     paste0("p ≈ ", round(pvalue2, 3)),
     pos = 4, cex = 1.2)
legend("topright",
       legend = c("Point Estimate"),
       lty = 2, lwd = 2, bty = "n",
       cex = 1.25)

```

## ANOVA and F-statistics

- When comparing **more than two groups**, we might be tempted to run multiple pairwise tests:
  - Group 1 vs. 2, Group 1 vs. 3, Group 2 vs. 3, etc.
  
- **Problem:** Multiple tests increase the chance of **false positives**
- **Solution**: Use a **single, holistic test** to compare all group means at once  
  
$$
H_0: \mu_1 = \mu_2 = \cdots = \mu_k
\quad\quad
H_A: \text{At least one mean differs}
$$

- ANOVA: tests whether there is evidence that **at least one mean differs**
  - **Between-group variation** (differences in group means)  
  - **Within-group variation** (random noise)
  - **Intuition:** If group differences are large *relative to* within-group variation, we reject the null.

- This produces the **F-statistic**

### ANOVA: Example 1

- Suppose a statistics department runs **three sections** (A, B, and C) of an introductory statistics course
  - We want to know whether there are **statistically significant differences** in the **average final exam scores**  

#### ANOVA: SSG

- The **sum of squares for the groups** (SSG) measures the differences between the group means $\bar{x}_i$ and the grand mean $\bar{x}$ like so:

$$
SSG = \sum_{i=1}^k n_i (\bar{x}_i - \bar{x})^2
$$

```{r}

# grades in each section
a <- c(90, 87, 88, 78)
b <- c(87, 85, 80)
c <- c(95, 93, 90, 88, 85)

# number of grades in each section
na <- length(a)
nb <- length(b)
nc <- length(c)

# average grades in each section
ua <- mean(a)
ub <- mean(b)
uc <- mean(c)

# grand mean
u <- mean(c(a, b, c))

# SSG
SSG <- na*(ua - u)^2 + nb*(ub - u)^2 + nc*(uc - u)^2
SSG

```

#### ANOVA: MSG

- The **mean square between groups** (MSG) is calculated using both the groups’ degrees of freedom $df_G$ and the sum of squares between groups SSG:

$$
MSG = \dfrac{1}{k - 1} SSG
$$

```{r}

k <- 3
df1 <- k - 1

MSG <- SSG / df1
MSG

```

#### ANOVA: SSE

- The **sum of squared errors** (SSE) is the sum of the squares of the differences between each sample’s observations with each respective sample mean
  - The calculation is simplified by first finding the variance $s_i^2$ of each sample:

$$
SSE = \sum_{i = 1}^k (n_i - 1) s_i^2
$$

```{r}

# standard errors for each group
sa <- sd(a)
sb <- sd(b)
sc <- sd(c)
sa; sb; sc

# SSE
SSE <- (na - 1)*sa^2 + (nb - 1)*sb^2 + (nc - 1)*sc^2
SSE

```

#### ANOVA: MSE

$$
MSE = \dfrac{1}{n - k} SSE
$$

```{r}

# MSE
n <- length(c(a, b, c))
MSE <- SSE / (n - k)
MSE

```

#### ANOVA: F-stat

- We now have comparable measures of variability both **between** the groups (MSG) and **within** the groups (MSE).
  - If variation among group means is simply random, F is typically close to 1

$$
F = \dfrac{MSG}{MSE}
$$
  
```{r}

# f-stat
f_stat <- MSG / MSE
f_stat

# generate f-dist
set.seed(1)
x <- rf(1e6, df1 = k - 1, df2 = n - k)
dx <- density(x)

# plot
par(mar = c(2, 4.5, 1, 1))
plot(dx$x, dx$y, type = "l", lwd = 2,
     ylab = "Density", cex.lab = 1.5, cex.axis = 1.25,
     xlim = c(1, 10), ylim = c(0, 0.4))
abline(v = f_stat, lty = 2, lwd = 2)
x_fill <- dx$x[dx$x >= f_stat]
y_fill <- dx$y[dx$x >= f_stat]
polygon(c(min(x_fill), x_fill, f_stat),
        c(0, y_fill, 0),
        col = rgb(0, 0, 1, 0.3), border = NA)
legend("topright",
       legend = c("f-stat"),
       lwd = 2, lty = 2, 
       cex = 1.25, bty = 'n')

# simulated p-value; FTR
1 - mean(ifelse(f_stat > x, 1, 0))

# estimated p-value; FTR
z <- data.frame(
  score = c(a, b, c),
  group = c(rep('a', na), rep('b', nb), rep('c', nc))
)
summary(aov(score ~ group, z))

```
